---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r prepare, message=FALSE, warning=FALSE} 
# change the below to the directory containing the downloaded activity.zip data file.
# setwd('D:/dokumenty/R WD/coursera courses/5 - reproducible research/RepData_PeerAssessment1')
require(data.table)
require(dplyr)
```

## Loading and preprocessing the data
```{r loaddata}
unzip("activity.zip")
activity = fread('activity.csv')

```

## What is mean total number of steps taken per day?
```{r stepsperday}
stepsbyday = 
    activity %>% group_by(date) %>% summarize(steps = sum(steps))

meansteps = round(mean(stepsbyday$steps, na.rm = TRUE)) 
meansteps
mediansteps = round(median(stepsbyday$steps, na.rm = TRUE))
mediansteps

hist(stepsbyday$steps, breaks = 14, 
     xlab = 'Steps per day', ylab = 'Day count', main = '', col = 'tomato')

abline(v = mediansteps, lwd = 2, lty = 2)

```

## What is the average daily activity pattern?

Getting average step count per intrval is straightforward enough...

```{r dailypattern-1}

stepsbyinterval = 
    activity[!is.na(activity$steps)] %>% 
    group_by(interval) %>% 
    summarize(steps = mean(steps))  %>%
    mutate(sequence = 1:288)
```

...but some manual processing is needed to to get the time intervals right. Note that coercing intervals to numbers leaves 'gaps' between 55 and 99 minute values, so instead a 1:288 sequence is used for plot X-axis but is labeled by hand with correct hours.

```{r dailypattern-2}

plot(stepsbyinterval$sequence, stepsbyinterval$steps, type = 'l', 
     main = 'Average step count in 5 minute inervals', 
     xlab = 'Time of day', ylab = 'Steps', 
     col = 'violetred3',
     axes = FALSE
     )

axis(1, labels = c('0:00','6:00','12:00','18:00','23:59'), at = c(0, 72, 144, 216, 288))
axis(2)
box()

maxstepsplot = stepsbyinterval$sequence[which.max(stepsbyinterval$steps)]
abline(v = maxstepsplot, lty = 2, col = 'violetred3')

maxstepsvalue = as.character(stepsbyinterval$interval[which.max(stepsbyinterval$steps)])
maxstepsvalue
```

The maximum number of steps in a five minute interval are on average made at `r maxstepsvalue`.

## Imputing missing values

Some of the days from the dataset have no reported data - they contain NA values instead. These are:

```{r NA-dates}
unique(activity[is.na(steps)][,date])
```

In each of these days, the NA values will be replaced by the mean of values from days for which the data is available. This will be done on a 5-minute interval basis. Then, mean and median of total steps per day are calculated in the same way as in the first step of the analysis. The results of these two calculations are compared to see if this strategy of filling in missing values introduces any changes to overall data distribution.

```{r missing}

fillednas = activity

for (i in 1:dim(fillednas)[1]){
    if (is.na(fillednas$steps[i])) {
        fillednas$steps[i] = stepsbyinterval$steps[stepsbyinterval$interval == fillednas$interval[i]]
    }
}

stepsbyday2 = 
    fillednas %>% group_by(date) %>% summarize(steps = sum(steps))

meansteps2 = round(mean(stepsbyday2$steps, na.rm = TRUE))
meansteps2
mediansteps2 = round(median(stepsbyday2$steps, na.rm = TRUE))
mediansteps2

if (identical(meansteps, meansteps2) & identical(mediansteps, mediansteps2)) {
    effect = 'no effect'
} else {
    effect = 'an effect'
}
```

Filling in the missing values has **`r effect`** on the mean and median compared to the original dataset. The histogram of data with filled NA values is as follows. 

```{r missing-2}
hist(stepsbyday2$steps, breaks = 14, 
     xlab = 'Steps per day', ylab = 'Day count', main = '', col = 'tomato')
abline(v = mediansteps2, lwd = 2, lty = 2)

```

## Are there differences in activity patterns between weekdays and weekends?
The dataset with NAs substituted by interval averages is split into weekday and weekend categories and interval averages for each group are calculated. A 1:288 sequence to use for plot X-axis is generated as before.

Then the two series are plotted on a single chart for comparison.

```{r differences}

differences = fillednas
differences$day = weekdays(strptime(differences$date, format='%Y-%M-%d'))
differences$weekend = 0

for (i in 1:dim(differences)[1]) {
    if (differences$day[i] == 'Saturday' | differences$day[i] == 'Sunday') {
        differences$weekend[i] = 1
    }
}

weekends = differences %>%
    select(interval, date, steps, weekend) %>%
    filter(weekend == 1) %>%
    group_by(interval) %>%
    summarize(steps = mean(steps)) %>%
    mutate(sequence = 1:288)

weekdays = differences %>%
    select(interval, date, steps, weekend) %>%
    filter(weekend == 0) %>%
    group_by(interval) %>%
    summarize(steps = mean(steps)) 

plot(weekends$sequence, weekends$steps, type = 'l', 
     main = 'Average step count in 5 minute inervals', 
     xlab = 'Time of day', ylab = 'Steps', 
     col = 'salmon',
     axes = FALSE)

axis(1, labels = c('0:00','6:00','12:00','18:00','23:59'), at = c(0, 72, 144, 216, 288))
axis(2)
box()

lines(weekends$sequence, weekdays$steps, col = 'navy')
legend('topright', legend = c('weekends','weekdays'), col = c('salmon','navy'), lwd = 1)
```

Looking at the graph it seems the walking patterns are quite similar between weekdays and weekends. Just to get some metric on this, let's calculate correclation.

```{r differences=2}
cor(weekends$steps, weekdays$steps)
```
